---
output:
  github_document:
    html_preview: false
    md_extensions: -smart
    pandoc_args: --wrap=none
---

# hypertext

<!-- badges: start -->

[![R-CMD-check](https://github.com/sigflux/hypertext/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sigflux/hypertext/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/hypertext)](https://CRAN.R-project.org/package=hypertext)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)

<!-- badges: end -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(hypertext)
```

html element construction for R.

`hypertext` provides a deterministic, framework-agnostic DSL for building
html nodes and rendering them to a string.

- dependency-free. zero package imports.
  lightweight is the right weight, as suggested in the
  [tinyverse](https://www.tinyverse.org/) philosophy.
- does **not** implement templating, dependency management, or
  widgets/framework integrations.
- heavily inspired by [{htmltools}](https://github.com/rstudio/htmltools).

## installation

install the stable version from CRAN:

```{r install_cran, eval = FALSE}
install.packages("hypertext")
```

or get the development version from github:

```{r install_github, eval = FALSE}
devtools::install_github("sigflux/hypertext")
```

## quick start

```{r quick_start}
page <- tag_list(
  doctype(),
  tags$html(
    tags$head(
      tags$title("hypertext")
    ),
    tags$body(
      tags$h1("Hello"),
      tags$p(
        class = c("lead", "mb-2"),
        "Server-side HTML."
      ),
      tags$input(
        type = "text",
        placeholder = "enter your nickname"
      ),
      tags$button(
        "Click"
      )
    )
  )
)

page
```

## rendering

`render()` takes a tag tree and returns a single HTML string:

```{r rendering}
x <- tags$p(
  class = "lead",
  "hello"
)

# `x` contains the tag tree
class(x)
x

# rendering produces an HTML string:
render(x)
```

you can `render()` directly to an html file by supplying
the `file` parameter:

```{r render_to_file, eval = FALSE}
page <- tag_list(
  doctype(),
  tags$html(
    tags$head(
      tags$title("hypertext")
    ),
    tags$body(
      tags$h1("Hello"),
      tags$p(
        class = c("lead", "mb-2"),
        "Server-side HTML."
      ),
      tags$input(
        type = "text",
        placeholder = "enter your nickname"
      ),
      tags$button(
        "Click"
      )
    )
  )
)

render(x = page, file = "index.html")
```

## tag lists

`tag_list()` groups sibling nodes without wrapping them in a parent element.

```{r tag_lists}
header <- tag_list(
  tags$h1("hello"),
  tags$p(
    class = "lead",
    "welcome aboard."
  )
)

# `header` is a "hypertext.tag.list" object:
class(header)

header
```

## raw html

`raw_html()` marks a string as pre-rendered HTML so that `render()` outputs
it verbatim, without escaping. useful for injecting inline scripts, styles,
SVG markup, or any content that is already valid HTML.

```{r raw_html}
tags$div(
  raw_html("<svg viewBox='0 0 100 100'><circle cx='50' cy='50' r='40'/></svg>")
)
```

`doctype()` is a convenience wrapper around `raw_html("<!DOCTYPE html>")`.

## custom tags

`tag()` creates elements for any tag name, including web components and
custom elements not in the built-in `tags` list. it takes 3 arguments:

- `tag_name`: name of the html element.
- `...`: attributes (named) & children (unnamed).
- `tag_type`: either "normal" (default) for the standard html
  elements, or "void" for the self-closing elements.

```{r custom_tag}
content <- tag(
  tag_name = "calcite-action-bar",
  layout = "horizontal"
)

content
```

nest them freely with each other and with built-in tags:

```{r nested_custom_tags}
page <- tags$div(
  class = "app",
  tag(
    tag_name = "calcite-shell",
    tag(
      tag_name = "calcite-shell-panel",
      slot = "panel-start",
      tag(
        tag_name = "calcite-action-bar",
        tag(
          tag_name = "calcite-action",
          text = "Layers",
          icon = "layers"
        ),
        tag(
          tag_name = "calcite-action",
          text = "Basemaps",
          icon = "basemap"
        )
      )
    ),
    tags$div(id = "map")
  )
)

page
```

for self-closing elements, set `tag_type = "void"`:

```{r void_custom_tag}
content <- tag(
  tag_name = "my-icon",
  name = "home",
  tag_type = "void"
)

content
```

## usage in frameworks

here are a couple of examples of how you can use hypertext in
R web frameworks.

### ambiorix

[ambiorix](https://ambiorix.dev/) is the perfect example of
  a web framework where you will find {hypertext} useful:

```{r ambiorix_usage, eval = FALSE}
library(ambiorix)
library(hypertext)

app <- Ambiorix$new(port = 3000L)

app$get("/", function(req, res) {
  html <- tags$h1("hello, world!") |>
    render()

  res$send(html)
})

app$get("/about", function(req, res) {
  html <- tag_list(
    tags$h1("about us"),
    tags$p(
      "minimal ",
      tags$strong("html construction"),
      " for R."
    )
  ) |>
    render()

  res$send(html)
})

app$get("/team", function(req, res) {
  teammates <- c("you", "me", "other")

  html <- tags$div(
    class = "team",
    tags$p("meet the team:"),
    tags$ul(
      lapply(teammates, tags$li)
    )
  ) |>
    render()

  res$send(html)
})

app$start()
```

### shiny

[shiny](https://shiny.posit.co/) already has {htmltools} tags
internally, so you do not need {hypertext} in your shiny apps,
but in case you do:

```{r shiny_usage, eval = FALSE}
library(shiny)
library(bslib)
library(hypertext)

# use `hypertext::tags` explicitly to avoid clashing with `shiny::tags`.
ht <- hypertext::tags

card <- function(title, body) {
  ht$div(
    class = "card mt-3",
    ht$div(
      class = "card-header",
      title
    ),
    ht$div(
      class = "card-body",
      ht$p(
        class = "card-text",
        body
      )
    )
  )
}

content <- ht$div(
  class = "container py-4",
  card("First card", "Some quick example text."),
  card("Second card", "Another body of text.")
) |>
  render()

ui <- page(
  theme = bs_theme(version = 5L),
  # hypertext renders an HTML string, so wrap in shiny::HTML()
  HTML(content)
)

server <- function(input, output, session) {}

shinyApp(ui, server)
```
