---
title: render
---

## render an HTML node tree to a character string

Converts a `hypertext.tag` object and all its descendants into an HTML string. Text children are escaped; nested `hypertext.tag` children are rendered recursively.

```r
render(x, ...)
render(x, file = "", write_mode = c("overwrite", "append"), ...)
```

## arguments

`x`

: `hypertext.tag` object, a string, or a list of these /// Required.

`...`

: Further arguments passed from or to other methods.

`file`

: String /// Optional. Path to file to print to.

`write_mode`

: String /// Optional. Either `"overwrite"` (default), which overwrites the contents of `file`, or `"append"`, which appends the HTML string to `file`.

## value

A single character string of HTML.

## details

When `file` is provided, the rendered HTML is written to the specified file via `cat()` and the HTML string is returned invisibly.

## examples

```{r}
library(hypertext)

page <- tags$html(
  tags$head(
    tags$title("Home")
  ),
  tags$body(
    tags$h1("Welcome")
  )
)

# return HTML as a string:
render(page)
```

```{r eval=FALSE}
# write to a file:
render(page, file = "index.html")
```
