---
title: "building components"
description: >
  how to wrap hypertext tags into reusable R functions,
  conditional children and composition.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(hypertext)
```

the core idea in hypertext is simple: tag functions are just R functions,
and R functions compose naturally. no need for a framework or a
special component system — a component is just a function that returns a
`hypertext.tag`.

with `hypertext`, you can use any css framework of your choice. but
for the examples that follow we'll use bootstrap classes for
demo purposes and because this site is built using `pkgdown` which
uses bootstrap by default.

## basic component

the simplest component is a function that wraps a tag with a fixed
structure, exposing only the parts that change as arguments.

```{r badge}
#' A Badge
#'
#' @param label String /// Required.
#'              Label to show on the badge.
#'
#' @param variant String /// Optional.
#'                Utility class. Must be one of:
#'                - "primary" (default)
#'                - "secondary"
#'                - "success"
#'                - "danger"
#'                - "warning"
#'                - "info"
#'                - "light"
#'                - "dark"
#'
#' @return `hypertext.tag`
#'
#' @export
badge <- function(
  label,
  variant = c(
    "primary",
    "secondary",
    "success",
    "danger",
    "warning",
    "info",
    "light",
    "dark"
  )
) {
  variant <- match.arg(arg = variant)

  tags$span(
    class = c(
      "badge",
      paste0("text-bg-", variant)
    ),
    label
  )
}

badge_a <- badge("new")
badge_b <- badge("deprecated", variant = "danger")
badge_c <- badge("stable", variant = "success")
```

```{r basic-components-badges, results = "asis"}
cat(
  render(badge_a),
  render(badge_b),
  render(badge_c)
)
```

the `variant` argument maps directly to a bootstrap utility class. the
caller controls the label and colour; the structure is fixed inside the
function.

## composition

components can call other components.

an `alert()` component can embed a `badge()` for the severity label,
keeping both independently reusable.

```{r alert}
#' An Alert
#'
#' @param message String /// Required.
#'                Message to show on the alert.
#'
#' @param type String /// Optional.
#'                Alert type. A utility class. Must be one of:
#'                - "info" (default)
#'                - "danger"
#'                - "warning"
#'
#' @param dismissible String /// Optional.
#'                 Should the alert be dismissible? Either:
#'                 - "no" (default)
#'                 - "yes"
#'
#' @return `hypertext.tag`
#'
#' @export
alert <- function(
  message,
  type = c("info", "danger", "warning"),
  dismissible = c("no", "yes")
) {
  type <- match.arg(arg = type)
  dismissible <- match.arg(arg = dismissible)

  label <- switch(
    type,
    info = badge(label = "info", variant = "primary"),
    warning = badge(label = "warning", variant = "warning"),
    danger = badge(label = "danger", variant = "danger"),
    badge(label = type)
  )

  classes <- c("alert", paste0("alert-", type))
  dismiss_btn <- NULL
  if (dismissible == "yes") {
    classes <- c(
      classes,
      "alert-dismissible fade show"
    )

    dismiss_btn <- tags$button(
      type = "button",
      class = "btn-close",
      `data-bs-dismiss` = "alert",
      `aria-label` = "Close"
    )
  }

  tags$div(
    class = classes,
    role = "alert",
    label,
    " ",
    message,
    dismiss_btn
  )
}

alert_a <- alert(
  message = "Your session will expire in 5 minutes.",
  type = "warning"
)

alert_b <- alert(
  message = "Record deleted successfully.",
  type = "danger"
)

alert_c <- alert(
  message = "This is some info you need to know.",
  dismissible = "yes"
)
```

```{r composition-alerts, results = "asis"}
cat(
  render(alert_a),
  render(alert_b),
  render(alert_c)
)
```

## Conditional children

Use `NULL` as a sentinel for optional children. `render()` drops `NULL`
values silently, so you can conditionally include children without
special-casing the rendering logic.

```{r card}
card <- function(
  title,
  body,
  footer = NULL
) {
  tags$div(
    class = "card",
    tags$div(
      class = "card-header",
      tags$strong(title)
    ),
    tags$div(
      class = "card-body",
      tags$p(
        class = "card-text",
        body
      )
    ),
    if (!is.null(footer)) {
      tags$div(
        class = "card-footer text-muted",
        footer
      )
    }
  )
}

# without footer
card_a <- card(
  title = "Title A",
  body = "Some body text."
)
# with footer
card_b <- card(
  title = "Title B",
  body = "Some body text.",
  footer = "Last updated: today"
)

render(card_a)
render(card_b)
```

```{r conditional-children-cards, results = "asis"}
cat(
render(card_a),
render(card_b)
)
```

When `footer` is `NULL` the `if` expression evaluates to `NULL`, which
hypertext simply ignores. No conditional rendering plumbing needed.

## List children

`lapply()` returns a plain list, and hypertext flattens plain lists of
children automatically. This means you can map over a vector to produce
a list of child nodes without any extra unwrapping.

```{r nav_bar}
nav_bar <- function(links) {
  # links: named character vector, names = labels, values = hrefs
  items <- lapply(
    X = names(links),
    FUN = function(label) {
      tags$li(
        class = "nav-item",
        tags$a(
          class = "nav-link",
          href = links[[label]],
          label
        )
      )
    }
  )

  tags$nav(
    tags$ul(
      class = "nav",
      items
    )
  )
}

site_links <- c(
  "Home" = "/",
  "Reference" = "/reference/",
  "Articles" = "/articles/"
)

my_nav <- nav_bar(site_links)
```

```{r list-chidren-nav, results = "asis"}
cat(
  render(my_nav)
)
```

## Worked example — profile card

Putting it together: a `profile_card()` component that combines an
image, a heading, a paragraph, and the `badge()` component from earlier.

```{r profile_card}
profile_card <- function(name, role, bio, avatar_url) {
  tags$div(
    class = "card",
    style = "max-width: 320px;",
    tags$img(
      src = avatar_url,
      class = "card-img-top",
      alt = paste("Photo of", name)
    ),
    tags$div(
      class = "card-body",
      tags$h5(class = "card-title", name),
      badge(role, variant = "secondary"),
      tags$p(class = "card-text mt-2", bio)
    )
  )
}

my_profile <- profile_card(
  name = "Ada Lovelace",
  role = "mathematician",
  bio = paste(
    "First to recognise that a computing machine had",
    "applications beyond pure calculation."
  ),
  avatar_url = "https://example.com/ada.jpg"
)
```

```{r worked-example-card, results = "asis"}
cat(
  render(my_profile)
)
```

## Summary

| Pattern | How |
|---|---|
| Fixed structure, variable content | Function argument → attribute or child |
| Optional children | Default to `NULL`; `if (!is.null(x))` guard |
| Composition | Call one component function inside another |
| Lists of children | `lapply()` → plain list → auto-flattened |
| Reuse across pages | Source the file or put components in a package |

A component library for a real project is just a collection of these
functions. There is no registration, no lifecycle, no virtual DOM —
just R functions that return tag objects.
